---
title: "Nich dynamics and ecological origins of migration - Analysis & Plots"
author: "Scott Yanco"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

# Initialization

Required libraries:
```{r message=FALSE}
#plotting
library(ggplot2)
library(ggthemes)
library(grid)
library(gridExtra)
# library(NineteenEightyR)
library(viridis)
library(extrafont)
library(RColorBrewer)

#data wrangling
library(lubridate)
library(tidyverse)

#spatial
library(sf)
library(sp)
library(rnaturalearth)
library(rnaturalearthdata)
library(raster)

#modeling
library(nlme)
library(glmmTMB)
library(ggeffects)
library(lattice)
library(tidymv)
library(mgcv)
library(brms)
library(fpp2)
```

Load data from previous script (rather than source data load and processing script which is very time consuming).

```{r}
load("../out/extract_2021-01-20.Rdata")
source("./functions.R")
```

# Figure S1 - Track Map

Make a map with all the tracks combined.  First we load in some background layers.

```{r}
#get world map of country boundaries 
world <- ne_countries(scale = "medium", returnclass = "sf")
#get US map of state boundaries
usa_states <- ne_states(returnclass = "sf", 
                        country = "united states of america")
#get MX map of state boundaries
mx_states <- ne_states(returnclass = "sf", country = "mexico")

#get US borders
us <- ne_countries(country = "united states of america", returnclass = "sp")
#get MX borders
mx <- ne_countries(country = "mexico", returnclass = "sp")
#get Guat. borders
guat <- ne_countries(country = "guatemala", returnclass = "sp")
#"sf" version of Guat.
guat_sf <- ne_countries(country = "guatemala", returnclass = "sf")

## Load Hypso Data ##
hypso <- raster("../data/gis/HYP_50M_SR_W/HYP_50M_SR_W/HYP_50M_SR_W.tif")

#crop hypso data to make processing faster
hypso <- crop(hypso, extent(matrix(c(-120, -90, 45, 10), byrow = T, nrow = 2)))

#get US hypso
hypso_us <- mask(hypso, us)
#get MX hypso
hypso_mx <- mask(hypso, mx)
#get Guat hypso
hypso_guat <- mask(hypso, guat)

#convert hypsos to spatial points data frames
hypso_us_spdf <- as(hypso_us, "SpatialPixelsDataFrame")
hypso_mx_spdf <- as(hypso_mx, "SpatialPixelsDataFrame")
hypso_guat_spdf <- as(hypso_guat, "SpatialPixelsDataFrame")

#convert hypsos to data.frames (for ggplot2)
hypso_us_df <- as.data.frame(hypso_us_spdf)
hypso_mx_df <- as.data.frame(hypso_mx_spdf)
hypso_guat_df <- as.data.frame(hypso_guat_spdf)

```


```{r}
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

(trackmap <- ggplot() +
  geom_tile(data = hypso_us_df, aes(x=x, y=y, fill = HYP_50M_SR_W), 
            alpha = .7) +
  geom_tile(data = hypso_mx_df, aes(x=x, y=y, fill = HYP_50M_SR_W), 
            alpha = .7) +
  geom_tile(data = hypso_guat_df, aes(x=x, y=y, fill = HYP_50M_SR_W), 
            alpha = .7) + 
  scale_fill_gradient2(low = "#01665E", mid = "#F5F5F5", high = "#8C510A", 
                       midpoint = 175) +
  geom_sf(data = world, fill = NA) +
  # geom_sf(data = mig_tracks_wint_sf[mig_tracks_wint_sf$season == "winter",], 
  #         inherit.aes = F, aes(group = bird_year, color  = bird_year), 
  #         size = 2) +
  geom_sf(data = mig_tracks_line, inherit.aes = F, 
          aes(group = bird_year, color = study_area), 
          size = 1, alpha = .75) +
  # scale_color_manual(values = cbbPalette) + 
  #ggtitle("Winter GPS Points Below 24 Degrees") +
  coord_sf(xlim = c(-120, -90), ylim = c(14, 45), expand = F) +
  theme_minimal() +
  theme(legend.position = "none", 
        # axis.text = element_text(family = "Arial Rounded MT Bold")
        ) +
  xlab("") +
  ylab(""))

ggsave("../figures/trackmap.png", trackmap)
```

# Experienced NT

## Models

```{r}
get_prior(bf(temp ~ 0 + season + (1|bird) + ar(time = date,gr = bird, cov = T), 
             sigma ~ 0 +season, family = gaussian()), data = temp_df)

fit_temp <- brm(bf(temp ~ 0 + season + (1|bird) + ar(time = date,gr = bird, cov = T), 
             sigma ~ 0 + season, family = gaussian()), data = temp_df, 
             chains = 4, iter = 10000, getOption("mc.cores", 3))
fit_temp
```

### Model performance checks:

Look at residuals and autocorrelation
```{r}
resid <- resid(fit_temp, type = "pearson")[, "Estimate"]
n <- length(resid)
resid[-1] <- resid[-1] - (resid[-n] * mean(as.matrix(fit_temp)[,
    "ar[1]"]))
fit <- fitted(fit_temp)[, "Estimate"]

ggplot() + geom_point(data = NULL, aes(y = resid, x = fit))
plot(acf(resid, lag = 40))
```


### Summarize

```{r}
sum_temp <- summary(fit_temp)

sum_temp$fixed

(temp_mod_df <- data.frame(param = c(rep("mu", 4), rep("sigma", 4)),
                          est = sum_temp$fixed[,1],
                          CI_l = sum_temp$fixed[,3],
                          CI_h = sum_temp$fixed[,4],
                          season = rep(c("summer", "fall", "winter", "spring"),
                                       2)))
```

## Plots
Create 4-color palette: 
```{r}
pal <- c("#4daf4a", "#e41a1c", "#984ea3", "#377eb8")
```


Reorder some factors for plotting
```{r}
temp_df$season <- fct_relevel(temp_df$season, 
                              c("spring", "summer", "fall", "winter"))
tempsummary$season <- fct_relevel(tempsummary$season, 
                                  c("spring", "summer", "fall", "winter"))
# tempmod$season <- fct_relevel(tempmod$season, 
#                               c("spring", "breeding", "fall", "winter"))
# tempvar$season <- fct_relevel(tempvar$season,
#                               c("spring", "summer", "fall", "winter"))
```

Make a common legend:
```{r fig.width=7, fig.height=0.5}
legend.plot <- ggplot(data = temp_df, 
                      aes(x = doy, y = temp, color = season)) +
  geom_point() +
  scale_color_manual(values = pal) +
  labs(color = "Season") +
  theme_few() +
  theme(legend.position = "left",
        legend.direction = "horizontal")

legend <- g_legend(legend.plot)
grid.arrange(legend)
```

Now plot panel A - raw temps.
```{r fig.width=7, fig.height=3}
#### Experienced temps ####
(rawtemps <- ggplot() +
  geom_point(data = temp_df, 
             aes(x = doy, y = temp, colour = season), size = 1.5) +
  scale_color_manual(values = pal) +
  annotate(geom="text", x=1, y=19, label="A", size = 7) +
  ylab(expression(paste("NT (",degree, "C)"))) +
  xlab("Day of Year") +
  labs(color = "Season") +
  theme_few() +
  theme(legend.position = "none"))
```

Plot panel B - estimated means from the model
```{r fig.width=3.5, fig.height=3}
temp_mod_df[temp_mod_df$param == "mu",]

(meantemps <- ggplot(temp_mod_df[temp_mod_df$param == "mu",]) +
  geom_point(aes(x = season, y = est, colour = season), size = 1.5) +
  geom_errorbar(aes(x = season, ymax = CI_h, ymin = CI_l, colour = season),
  width =.5, size = .5) +
  annotate(geom="text", x=4.2, y=14.7, label="B", size = 7) +
  # scale_x_discrete(labels = c("Spring", "Summer", "Fall", "Winter")) +
  scale_colour_manual(values = pal) +
  ylim(0, 15) +
  ylab(expression(paste("Mean NT (",degree, "C)"))) +
  xlab("Season") + 
  theme_few() +
  theme(legend.position = "none"))
```

```{r fig.width=3.5, fig.height=3}
temp_mod_df[temp_mod_df$param == "sigma",]
(SDtemps <- ggplot(temp_mod_df[temp_mod_df$param == "sigma",]
) +
  geom_point(aes(x = season, y = est, color = season), size = 1.5) +
  geom_errorbar(aes(x = season, ymin = CI_l, ymax = CI_h, color = season),
                width = .5, size = .5) +
  scale_colour_manual(values = pal) +
  scale_x_discrete(labels = c("Spring", "Summer", "Fall", "Winter")) +
  annotate(geom="text", x=4.2, y=1.8, label="C", size = 7) +
  ylab(expression(paste("Std. Dev. NT (",degree, "C)"))) +
  xlab("Season") + 
  theme_few() +
  theme(legend.position = "none"))

```

### Arrange the plots into single figure

```{r fig.height=10, fig.width=7}
gl <- list(rawtemps, legend, meantemps, SDtemps)

ntfig <- arrangeGrob(
  grobs = gl,
  heights = unit(c(3,.5, 3), c("in", "in", "in")),
  #  widths = c(2, 1, 1),
  layout_matrix = rbind(c(1, 1),
                        c(2, 2),
                        c(3, 4))
)

ggsave("../figures/nT_fig.pdf", ntfig)
ggsave("../figures/nT_fig.png", ntfig)

```


# Experienced NDVI

## Models

Build the linear model that includes:  
- Random intercept by bird;
- First-order auto-regressive covariance structure
- Heterogeneous variance by season
- AND models NDVI as beta distributed

Need to make a factior date for this AR1
```{r}
ndvi_df$date_factor <- factor(ndvi_df$date)
ndvi_df$bird <- factor(ndvi_df$bird)
```


First fit the models.
```{r}
get_prior(bf(ndvi ~ season + (1|bird) + ar(time = date,gr = bird, cov = T), 
             phi ~ season, family = Beta()), data = ndvi_df)

fit_ndvi <- brm(bf(ndvi ~ season + (1|bird) + ar(time = date, gr = bird, cov = T), 
             phi ~ season, family = Beta()), data = ndvi_df, 
             chains = 4, iter = 1000, cores = getOption("mc.cores", 3),
             control = list(adapt_delta = 0.95))
fit_ndvi
```
### Model performance checks

```{r}
resid <- resid(fit_ndvi, type = "pearson")[, "Estimate"]
n <- length(resid)
resid[-1] <- resid[-1] - (resid[-n] * mean(as.matrix(fit_ndvi)[,
    "ar[1]"]))
fit <- fitted(fit_ndvi)[, "Estimate"]

ggplot() + geom_point(data = NULL, aes(y = resid, x = fit))
plot(acf(resid, lag = 40))
```


### Summarize

```{r}
sum_ndvi <- summary(fit_ndvi)

sum_ndvi$fixed

(ndvi_mod_df <- data.frame(param = c("mu", "phi", "mu", "mu", "mu", 
                                     "phi", "phi", "phi"),
                          est = sum_ndvi$fixed[,1],
                          CI_l = sum_ndvi$fixed[,3],
                          CI_h = sum_ndvi$fixed[,4],
                          season = c("summer", "summer", "fall", "winter", 
                                         "spring","fall", "winter", "spring"),
                          est_2 = NA,
                          CI_l_2 = NA,
                          CI_h_2 = NA))

#hacky code to deal with offsets - remove once model refitted through intercept
for (i in 3:nrow(ndvi_mod_df)) {
  if(ndvi_mod_df$param[i] == "mu"){
    ndvi_mod_df$est_2[i] <- ndvi_mod_df$est[i] + ndvi_mod_df$est[1]
    ndvi_mod_df$CI_l_2[i] <- ndvi_mod_df$CI_l[i] + ndvi_mod_df$CI_l[1]
    ndvi_mod_df$CI_h_2[i] <- ndvi_mod_df$CI_h[i] + ndvi_mod_df$CI_h[1]

  }  else {
    ndvi_mod_df$est_2[i] <- ndvi_mod_df$est[i] + ndvi_mod_df$est[2]
    ndvi_mod_df$CI_l_2[i] <- ndvi_mod_df$CI_l[i] + ndvi_mod_df$CI_l[2]
    ndvi_mod_df$CI_h_2[i] <- ndvi_mod_df$CI_h[i] + ndvi_mod_df$CI_h[2]
  }
}

ndvi_mod_df$est_2[1] <- ndvi_mod_df$est[1]
ndvi_mod_df$est_2[2] <- ndvi_mod_df$est[2]

ndvi_mod_df$CI_l_2[1] <- ndvi_mod_df$CI_l[1]
ndvi_mod_df$CI_l_2[2] <- ndvi_mod_df$CI_l[2]

ndvi_mod_df$CI_h_2[1] <- ndvi_mod_df$CI_h[1]
ndvi_mod_df$CI_h_2[2] <- ndvi_mod_df$CI_h[2]

```

## Plots
Reorder some factors for plotting
```{r}
ndvi_mod_df$season <- fct_relevel(ndvi_mod_df$season, 
                              c("spring", "summer", "fall", "winter"))
ndvi_df$season <- fct_relevel(ndvi_df$season, 
                                  c("spring", "summer", "fall", "winter"))
# ndvimod$season <- fct_relevel(ndvimod$season, 
#                               c("spring", "breeding", "fall", "winter"))
# ndvivar$season <- fct_relevel(ndvivar$season,
#                               c("spring", "summer", "fall", "winter"))
```

Now plot panel A - raw NDVI.
```{r fig.width=7, fig.height=3}
#### Experienced temps ####
(rawndvi <- ggplot() +
  geom_point(data = ndvi_df, 
             aes(x = doy, y = ndvi, colour = season),
             size = 1.5) +
  annotate(geom="text", x=1, y=.95, label="A", size = 7) +
  scale_color_manual(values = pal) +
  ylim(0,1) +
  ylab("NDVI") +
  xlab("Day of Year") +
  labs(color = "Season") +
  theme_few() +
  theme(legend.position = "none"))
```

Plot panel B - estimated means from the model
# CHANGE EST_2 TO EST AFTER REFITTING MODEL THROUGH INTERECPT
```{r fig.width=3.5, fig.height=3}
(meanndvi <- ggplot(ndvi_mod_df[ndvi_mod_df$param == "mu",]) +
  geom_point(aes(x = season, y = est_2, colour = season), size = 1.5) +
  geom_errorbar(aes(x = season, ymax = CI_h_2, ymin = CI_l_2, colour = season), 
                width =.5, size = .5) +
  scale_x_discrete(labels = c("Spring", "Summer", "Fall", "Winter")) +
  scale_colour_manual(values = pal) +
  annotate(geom="text", x=4.2, y=.95, label="B", size = 7) +
  # ylim(0, 1) +
  ylab("Mean NDVI") +
  xlab("Season") + 
  theme_few() +
  theme(legend.position = "none"))
```

Plot Panel C - strata-specific variances
```{r fig.width=3.5, fig.height=3}
(SDndvi <- ggplot(ndvi_mod_df[ndvi_mod_df$param == "phi",]) +
  geom_point(aes(x = season, y = est_2, color = season), size = 1.5) +
  geom_errorbar(aes(x = season, ymin = CI_l_2, ymax = CI_h_2, color = season),
                width = .5, size = .5) +
  scale_colour_manual(values = pal) +
  scale_x_discrete(labels = c("Spring", "Summer", "Fall", "Winter")) +
  annotate(geom="text", x=4.2, y=.5, label="C", size = 7) +
  ylab("Std. Dev. NDVI") +
  xlab("Season") + 
  theme_few() +
  theme(legend.position = "none"))

```

### Arrange into single
```{r fig.height=10, fig.width=7}
gl2 <- list(rawndvi, legend, meanndvi, SDndvi)

ntfig <- arrangeGrob(
  grobs = gl2,
  heights = unit(c(3,.5, 3), c("in", "in", "in")),
  #  widths = c(2, 1, 1),
  layout_matrix = rbind(c(1, 1),
                        c(2, 2),
                        c(3, 4))
)

ggsave("../figures/ndvi_fig.pdf", plot = ntfig)
ggsave("../figures/ndvi_fig.png", ntfig)
```

## Save Model Objects 
```{r}
save(fit_ndvi, fit_temp, file = paste0("../out/model_fits_", Sys.Date(), ".Rdata"))
```


#Counterfactual Breeding
## Counterfactual NT

```{r}
hypo_temp_df$season <- fct_relevel(hypo_temp_df$season, 
                                   c("spring", "summer", "fall", "winter"))
merge_temps$season <- fct_relevel(merge_temps$season, 
                                  c("spring", "summer", "fall", "winter"))
paired_temps$season <- fct_relevel(paired_temps$season, 
                                  c("spring", "summer", "fall", "winter"))
```


Make a plot of raw temps with raw hypo temps.
```{r fig.width=3.5, fig.height=3}
(rawhypoNT <- ggplot() +
   geom_point(data = temp_df, 
              aes(x = doy, y = temp), color = "lightgray", alpha = .7, 
              shape = 17) +
   geom_point(data = hypo_temp_df,
              aes(x = doy, y = temp, color = season)) +
   scale_color_manual(values = pal) +
   ylab(expression(paste("NT (",degree, "C)"))) +
   xlab("Day of Year") +
   theme_few() +
   xlim(0, 366) +
   ylim(-15, 30) +
   theme(legend.position = "none"))
```
```{r}
ggsave("../figures/hyp_summer/rawNT.png", rawhypoNT, width = 3.5, height = 3)
```

```{r fig.width=3.5, fig.height=3}
(meanhypoNT <- ggplot() +
   
   #real points and error bars
   geom_errorbar(data = paired_temps[paired_temps$type == "real",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, group = bird), 
                 width = .5, size = .5, color = "grey44",
                 position=position_dodge(width=0.9)) +
   geom_point(data = paired_temps[paired_temps$type == "real",],
              aes(x = season, y = mu, group = bird),
              shape = 17, size = 1.5, color = "grey44",
              position=position_dodge(width=0.9)) +
   
   #simulated points and bars
   geom_errorbar(data = paired_temps[paired_temps$type == "hypo",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, 
                     color = season, group = bird),
                 width = .5, size = .5, position=position_dodge(width=0.9)) +
   geom_point(data = paired_temps[paired_temps$type == "hypo",],
              aes(x = season, y = mu, color = season, group = bird),
              size = 1.5, position=position_dodge(width=0.9)) +
   
   scale_color_manual(values = pal, 
                      name = "Season") +
   # scale_x_discrete(labels = c("Spring", "Summer", "Fall")) +
   theme_few()+
   theme(axis.ticks.x = element_blank(),
         legend.position = "none",
         strip.text = element_blank()) +
   ylab(expression(paste("NT (",degree, "C)"))) +
   xlab("Season") +
   facet_grid(cols = vars(season), scales = "free"))
```
Get mean pairwise difference:
```{r}
mean_pair_NT <- paired_temps %>% 
  pivot_wider(id_cols = bird, names_from = c(season, type), values_from = mu)

#spring
t.test(mean_pair_NT$spring_hypo, mean_pair_NT$spring_real, paired = T)

#summer
t.test(mean_pair_NT$summer_hypo, mean_pair_NT$summer_real, paired = T)

#fall
t.test(mean_pair_NT$fall_hypo, mean_pair_NT$fall_real, paired = T)
```

```{r}
ggsave("../figures/hyp_summer/muNT.png", meanhypoNT, width = 3.5, height = 3)
```

## Counterfactual NDVI

```{r}
ndvi_hypo_df$season <- fct_relevel(ndvi_hypo_df$season, 
                                   c("spring", "summer", "fall", "winter"))
merge_ndvi$season <- fct_relevel(merge_ndvi$season, 
                                 c("spring", "summer", "fall", "winter"))
paired_ndvi$season <- fct_relevel(paired_ndvi$season, 
                                 c("spring", "summer", "fall", "winter"))
```


Make a plot of raw temps with raw hypo temps.
```{r fig.width=3.5, fig.height=3}
(rawhypoNDVI <- ggplot() +
   geom_point(data = ndvi_df, 
              aes(x = doy, y = ndvi), color = "lightgray", alpha = .7, 
              shape = 17) +
   geom_point(data = ndvi_hypo_df[ndvi_hypo_df$season != "winter",],
              aes(x = doy, y = ndvi, color = season)) +
   scale_color_manual(values = pal) +
   ylab("NDVI") +
   xlab("Day of Year") +
   theme_few() +
   xlim(0, 366) +
   ylim(0,1) +
   theme(legend.position = "none"))
```
```{r}
ggsave("../figures/hyp_summer/rawNDVI.png", rawhypoNDVI, width = 3.5, height = 3)
```


```{r fig.width=3.5, fig.height=3}
season.names <- c("Spring", "Summer", "Fall")
names(season.names) <- c("spring", "summer", "fall")

(meanhypoNDVI <- ggplot() +
    
    #real points and error bars
    geom_errorbar(data = paired_ndvi[paired_ndvi$type == "real",],
                  aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, group = bird), 
                  width = .3, size = 1, color = "black", alpha = 0.4,
                  position=position_dodge(width=0.9)) +
    geom_point(data = paired_ndvi[paired_ndvi$type == "real",],
               aes(x = season, y = mu, group = bird),
               shape = 17, size = 1.5, color = "black", alpha = 0.4,
               position=position_dodge(width=0.9)) +
    
    #simulated points and bars
    geom_errorbar(data = paired_ndvi[paired_ndvi$type == "hypo",],
                  aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, 
                      color = season, group = bird),
                  width = .3, size = 1, position=position_dodge(width=0.9)) +
    geom_point(data = paired_ndvi[paired_ndvi$type == "hypo",],
               aes(x = season, y = mu, color = season, group = bird),
               size = 1.5, position=position_dodge(width=0.9)) +
    
    scale_color_manual(values = pal, 
                       name = "Season") +
    # scale_x_discrete(labels = c("Spring", "Summer", "Fall")) +
    theme_few()+
    theme(axis.ticks.x = element_blank(),
          legend.position = "none", strip.placement = "bottom", 
          plot.title = element_text(hjust=0),
          strip.text = element_blank()) +
    ylab(expression(paste("NT (",degree, "C)"))) +
    xlab("Season") +
    ylab("NDVI") +
    facet_grid(cols = vars(season), scales = "free"))
```
```{r}
ggsave("../figures/hyp_summer/muNDVI.png", meanhypoNDVI, width = 3.5, height = 3)
```

Get mean pairwise difference:
```{r}
mean_pair_ndvi <- paired_ndvi %>% 
  pivot_wider(id_cols = bird, names_from = c(season, type), values_from = mu)

#spring
t.test(mean_pair_ndvi$spring_hypo, mean_pair_ndvi$spring_real, paired = T)

#summer
t.test(mean_pair_ndvi$summer_hypo, mean_pair_ndvi$summer_real, paired = T)

#fall
t.test(mean_pair_ndvi$fall_hypo, mean_pair_ndvi$fall_real, paired = T)
```

## Common legend
Extract common legend.

```{r fig.width=7, fig.height = 0.5}
hypSummLeg <- ggplot(merge_temps) +
  geom_point(data = merge_temps, aes(x = season, color = season, y = mu,
                                     alpha = type, shape = type)) +
  geom_point(data = merge_temps_wint, aes(x = season, color = season, y = mu,
                                          alpha = type, shape = type)) + 
  scale_color_manual(values = pal, 
                     labels = c("Spring", "Summer", "Fall", "Winter"),
                     name = "Season") +
  scale_x_discrete(labels = c("Spring", "Summer", "Fall", "Winter")) +
  scale_alpha_manual(values = c(1, 0.3), 
                     labels = c("Simulated", "Experienced"),
                     name = "Data Type") +
  scale_shape_discrete(labels = c("Simulated", "Experienced"),
                       name = "Data Type") +
  theme_few() +
  theme(legend.box = "horizontal", legend.position = "bottom")


legend <- g_legend(hypSummLeg)
grid.arrange(legend)
```
```{r}
ggsave("../figures/hyp_summer/legend.png", legend, width = 7, height = 0.5)
```

#Counterfactual Winter
```{r}
pal2 <- c("#4daf4a", "#984ea3", "#377eb8")
```

## Counterfactual NT
```{r}
hypo_temp_wint_df$season <- fct_relevel(hypo_temp_wint_df$season, 
                                        c("spring", "summer", "fall", "winter"))
merge_temps_wint$season <- fct_relevel(merge_temps_wint$season, 
                                       c("spring", "summer", "fall", "winter"))
paired_temps_wint$season <- fct_relevel(paired_temps_wint$season, 
                                       c("spring", "summer", "fall", "winter"))
```

Make a plot of raw temps with raw hypo temps.
```{r}
(rawhypoNTwint <- ggplot() +
   geom_point(data = temp_df, 
              aes(x = doy, y = temp), color = "lightgray", alpha = .7, 
              shape = 17) +
   geom_point(data = hypo_temp_wint_df[hypo_temp_wint_df$season != "summer",],
              aes(x = doy, y = temp, color = season)) +
   scale_color_manual(values = pal2)+
   ylab(expression(paste("NT (",degree, "C)"))) +
   xlab("Day of Year") +
   theme_few() +
   xlim(0, 366) +
   theme(legend.position = "none"))
```

```{r}
ggsave("../figures/hyp_winter/rawNT.png", rawhypoNTwint, width = 3.5, height = 3)
```

```{r}
(meanhypoNTwint <- ggplot() +
   
   #real points and error bars
   geom_errorbar(data = paired_temps_wint[paired_temps_wint$type == "real",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, group = bird), 
                 width = .5, size = .5, color = "grey44",
                 position=position_dodge(width=0.9)) +
   geom_point(data = paired_temps_wint[paired_temps_wint$type == "real",],
              aes(x = season, y = mu, group = bird),
              shape = 17, size = 1.5, color = "grey44",
              position=position_dodge(width=0.9)) +
   
   #simulated points and bars
   geom_errorbar(data = paired_temps_wint[paired_temps_wint$type == "hypo",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, 
                     color = season, group = bird),
                 width = .5, size = .5, position=position_dodge(width=0.9)) +
   geom_point(data = paired_temps_wint[paired_temps_wint$type == "hypo",],
              aes(x = season, y = mu, color = season, group = bird),
              size = 1.5, position=position_dodge(width=0.9)) +
   
   scale_color_manual(values = pal2)+
   #scale_x_discrete(labels = c("Spring", "Summer", "Fall")) +
   # scale_alpha_manual(values = c(1, 0.3), 
   #                    labels = c("Simulated", "Experienced"),
   #                    name = "Data Type") +
   # scale_shape_discrete(labels = c("Simulated", "Experienced"),
   #                      name = "Data Type") +
   theme_few()+
   theme(axis.ticks.x = element_blank(),
         legend.position = "none", strip.placement = "bottom", 
         plot.title = element_text(hjust=0),
         strip.text = element_blank()) +
   ylab(expression(paste("NT (",degree, "C)"))) +
   xlab("Season") +
   ylab(expression(paste("NT (",degree, "C)"))) +
   
   facet_grid(cols = vars(season), scales = "free"))
```
```{r}
ggsave("../figures/hyp_winter/meanNT.png", meanhypoNTwint, width = 3.5, height = 3)
```

Get mean pairwise difference:
```{r}
mean_pair_nt_wint <- paired_temps_wint %>% 
  pivot_wider(id_cols = bird, names_from = c(season, type), values_from = mu)

#spring
t.test(mean_pair_nt_wint$spring_hypo, mean_pair_nt_wint$spring_real, paired = T)

#winter
t.test(mean_pair_nt_wint$winter_hypo, mean_pair_nt_wint$winter_real, paired = T)

#fall
t.test(mean_pair_nt_wint$fall_hypo, mean_pair_nt_wint$fall_real, paired = T)
```

## Counterfactual NDVI
```{r}
ndvi_hypo_wint_df$season <- fct_relevel(ndvi_hypo_wint_df$season, 
                                        c("spring", "summer", "fall", "winter"))
merge_ndvi_wint$season <- fct_relevel(merge_ndvi_wint$season, 
                                      c("spring", "summer", "fall", "winter"))
```

Make a plot of raw temps with raw hypo temps.
```{r}
(rawhypoNDVIwint <- ggplot() +
   geom_point(data = ndvi_df, 
              aes(x = doy, y = ndvi), color = "lightgray", alpha = .7, 
              shape = 17) +
   geom_point(data = ndvi_hypo_wint_df[ndvi_hypo_wint_df$season != "summer",],
              aes(x = doy, y = temp, color = season)) +
   scale_color_manual(values = c("#E41A1C", "#4DAF4A", "#984EA3"))+
   ylab("NDVI") +
   xlab("Day of Year") +
   theme_few() +
   xlim(0, 366) +
   theme(legend.position = "none"))
```

```{r}
ggsave("../figures/hyp_winter/rawNNDVI.png", rawhypoNDVIwint, width = 3.5, height = 3)
```

```{r}
(meanhypoNDVIwint <- ggplot() +
   
   #real points and error bars
   geom_errorbar(data = merge_ndvi_wint[merge_ndvi_wint$type == "real",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, group = bird), 
                 width = .5, size = .5, color = "grey44",
                 position=position_dodge(width=0.9)) +
   geom_point(data = merge_ndvi_wint[merge_ndvi_wint$type == "real",],
              aes(x = season, y = mu, group = bird),
              shape = 17, size = 1.5, color = "grey44",
              position=position_dodge(width=0.9)) +
   
   #simulated points and bars
   geom_errorbar(data = merge_ndvi_wint[merge_ndvi_wint$type == "hypo",],
                 aes(x = season, ymax = mu.CI.h, ymin = mu.CI.l, 
                     color = season, group = bird),
                 width = .5, size = .5, position=position_dodge(width=0.9)) +
   geom_point(data = merge_ndvi_wint[merge_ndvi_wint$type == "hypo",],
              aes(x = season, y = mu, color = season, group = bird),
              size = 1.5, position=position_dodge(width=0.9)) +
   
   scale_color_manual(values = c("#E41A1C", "#4DAF4A", "#984EA3"))+
   theme_few()+
   theme(axis.ticks.x = element_blank(),
         legend.position = "none", strip.placement = "bottom", 
         plot.title = element_text(hjust=0),
         strip.text = element_blank()) +
   ylab("NDVI") +
   xlab("Season") +
   facet_grid(cols = vars(season), scales = "free"))
```
```{r}
ggsave("../figures/hyp_winter/meanNDVI.png", meanhypoNDVIwint, width = 3.5, height = 3)
```


## Common legend
Extract common legend.

```{r fig.width=7, fig.height = 0.5}
hypWintLeg <- ggplot(merge_temps) +
  # geom_point(data = merge_temps, aes(x = season, color = season, y = mu,
  #                                    alpha = type, shape = type)) +
  geom_point(data = merge_temps_wint, aes(x = season, color = season, y = mu,
                                          alpha = type, shape = type)) + 
  scale_color_manual(values = pal2, 
                     labels = c("Spring", "Fall", "Winter"),
                     name = "Season") +
  scale_x_discrete(labels = c("Spring", "Fall", "Winter")) +
  scale_alpha_manual(values = c(1, 0.3), 
                     labels = c("Simulated", "Experienced"),
                     name = "Data Type") +
  scale_shape_discrete(labels = c("Simulated", "Experienced"),
                       name = "Data Type") +
  theme_few() +
  theme(legend.box = "horizontal", legend.position = "bottom")


legend <- g_legend(hypWintLeg)
grid.arrange(legend)
ggsave("../figures/hyp_winter/legend.png", hypWintLeg, width = 7, height = 0.5)
```

# Breeding Season Transitions

## Min temp

First build GAMs of minimum daily  temps
```{r}
#build GAM of temp for local weather stations (Platoro-HC, Florissant-MEF)
fm_flor <- gam(TMIN ~ s(doy), data = flor_tmin)
fm_plat <- gam(TMIN ~ s(doy), data = platoro)

summary(fm_flor)
summary(fm_plat)
#extract fitted values of gams
pred_flor <- predict_gam(fm_flor, length_out = 366)
pred_plat <- predict_gam(fm_plat, length_out = 366)
```

```{r}
#extract fitted temp value that matches median departure/arrival dates

#HC
(hc_dep <- pred_plat$fit[pred_plat$doy == median(na.omit(hc_time$departure))])
(hc_arr <- pred_plat$fit[pred_plat$doy == median(na.omit(hc_time$arrival))])
(hc_dif <- hc_dep-hc_arr)
```
```{r}
(mef_dep <- pred_flor$fit[pred_flor$doy == median(na.omit(mef_time$departure))])
(mef_arr <- pred_flor$fit[pred_flor$doy == median(na.omit(mef_time$arrival))])
(mef_dif <- mef_dep-mef_arr)

max(pred_flor$fit) - min(pred_flor$fit)
max(pred_plat$fit) - min(pred_plat$fit)
max(pred_flor$fit)
max(pred_plat$fit)

```

Plot models with transition dates

```{r fig.width=3.5, fig.height=2}
(#plot HC min temps (GAM) with departures and arrivals
hc_plot <- ggplot() +
  geom_ribbon(data = pred_plat, 
              aes(ymin = (fit-se.fit), ymax = (fit+se.fit), x = doy), 
              color = "gray", alpha = 0.25) +
  geom_line(data = pred_plat, aes(x = doy, y = fit), inherit.aes = F) +
  geom_vline(data = hc_time, aes(xintercept = departure), color = "red") +
  geom_vline(data = hc_time,
             aes(xintercept = median(na.omit(departure))), size = 1) +
  geom_vline(data = hc_time, aes(xintercept = arrival), color = "red") +
  geom_vline(data = hc_time, 
             aes(xintercept = median(na.omit(arrival))), size = 1) +
  geom_hline(yintercept = hc_dep, linetype = "twodash", color = "blue") +
  geom_hline(yintercept = hc_arr, linetype = "twodash", color = "purple") +
  xlab("Day of Year") +
  ylab("Temp (C)") +
  ylim(-20,10) +
  #geom_text(aes(x = 55, y = 8, label = "Conejos County, CO")) +
  theme_few())

#plot MEF min temps (GAM) with departures and arrivals
(mef_plot <- ggplot() +
  geom_ribbon(data = pred_flor, 
              aes(ymin = (fit-se.fit), ymax = (fit+se.fit), x = doy), 
              color = "gray", alpha = 0.25) +
  geom_line(data = pred_flor, aes(x = doy, y = fit), inherit.aes = F) +
  geom_vline(data = mef_time, aes(xintercept = departure), color = "red") +
  geom_vline(data = mef_time,
             aes(xintercept = median(departure)), size = 1) +
  geom_vline(data = mef_time, aes(xintercept = arrival), color = "red") +
  geom_vline(data = mef_time,
             aes(xintercept = median(na.omit(arrival))), size = 1) +
  geom_hline(yintercept = mef_dep, linetype = "twodash", color = "blue") +
  geom_hline(yintercept = mef_arr, linetype = "twodash", color = "purple") +
  xlab("Day of Year") +
  ylab("Temp (C)") +
  ylim(-20,10) +
  #geom_text(aes(x = 50, y = 8, label = "Teller County, CO")) +
  theme_few())

# #combine plots
# CO_temps <- arrangeGrob(mef_plot, hc_plot)
# ggsave("breed_temps.png", CO_temps)
```

## NDVI

First build GAMs of NDVI
```{r}
#build GAM
fm_HC_ndvi <- gam(ndvi ~ s(doy), data = HC_ndvi_ts)
fm_MEF_ndvi <- gam(ndvi ~ s(doy), data = MEF_ndvi_ts)


#extract fitted values of gams
pred_HC_ndvi <- predict_gam(fm_HC_ndvi, values = list(doy = 1:366))
pred_MEF_ndvi <- predict_gam(fm_MEF_ndvi, values = list(doy = 1:366))


```

```{r}
#extract fitted temp value that matches median departure/arrival dates

#HC
(hc_dep_ndvi <- pred_HC_ndvi$fit[pred_HC_ndvi$doy == 
                                   median(na.omit(hc_time$departure))])
(hc_arr_ndvi <- pred_HC_ndvi$fit[pred_HC_ndvi$doy == 
                                   median(na.omit(hc_time$arrival))])
(hc_dif_ndvi <- hc_dep_ndvi-hc_arr_ndvi)

#MEF
(mef_dep_ndvi <- pred_MEF_ndvi$fit[pred_MEF_ndvi$doy ==
                                     median(na.omit(mef_time$departure))])
(mef_arr_ndvi <- pred_MEF_ndvi$fit[pred_MEF_ndvi$doy == 
                                     median(na.omit(mef_time$arrival))])
(mef_dif_ndvi <- mef_dep_ndvi-mef_arr_ndvi)

#total range of NDVI across the year
max(pred_MEF_ndvi$fit) - min(pred_MEF_ndvi$fit)
max(pred_HC_ndvi$fit) - min(pred_HC_ndvi$fit)
max(pred_MEF_ndvi$fit)
max(pred_HC_ndvi$fit)
```

Plot it!
```{r fig.width=3.5, fig.height=2}
#plot HC NDVI (GAM) with departures and arrivals
(hc_ndvi_plot <- ggplot() +
  geom_ribbon(data = pred_HC_ndvi, 
              aes(ymin = (fit-se.fit), ymax = (fit+se.fit), x = doy), 
              color = "gray", alpha = 0.25) +
  geom_line(data = pred_HC_ndvi, aes(x = doy, y = fit), inherit.aes = F) +
  geom_vline(data = hc_time, aes(xintercept = departure), color = "red") +
  geom_vline(data = hc_time,
             aes(xintercept = median(na.omit(departure))), size = 1) +
  geom_vline(data = hc_time, aes(xintercept = arrival), color = "red") +
  geom_vline(data = hc_time, 
             aes(xintercept = median(na.omit(arrival))), size = 1) +
  geom_hline(yintercept = hc_dep_ndvi, linetype = "twodash", color = "blue") +
  geom_hline(yintercept = hc_arr_ndvi, linetype = "twodash", color = "purple") +
  xlab("Day of Year") +
  ylab("NDVI") +
  ylim(0,0.6) +
  #geom_text(aes(x = 55, y = 8, label = "Conejos County, CO")) +
  theme_few())
```

```{r fig.width=3.5, fig.height=2}
#plot MEF NDVI (GAM) with departures and arrivals
(mef_ndvi_plot <- ggplot() +
  geom_ribbon(data = pred_MEF_ndvi, 
              aes(ymin = (fit-se.fit), ymax = (fit+se.fit), x = doy), 
              color = "gray", alpha = 0.25) +
  geom_line(data = pred_MEF_ndvi, aes(x = doy, y = fit), inherit.aes = F) +
  geom_vline(data = mef_time, aes(xintercept = departure), color = "red") +
  geom_vline(data = mef_time,
             aes(xintercept = median(na.omit(departure))), size = 1) +
  geom_vline(data = mef_time, aes(xintercept = arrival), color = "red") +
  geom_vline(data = mef_time, 
             aes(xintercept = median(na.omit(arrival))), size = 1) +
  geom_hline(yintercept = mef_dep_ndvi, linetype = "twodash", color = "blue") +
  geom_hline(yintercept = mef_arr_ndvi, linetype = "twodash", color = "purple") +
  xlab("Day of Year") +
  ylab("NDVI") +
  ylim(0,0.6) +
  #geom_text(aes(x = 55, y = 8, label = "Conejos County, CO")) +
  theme_few())

```

## Combine Plots

```{r fig.width=8, fig.height=5}
gl_phen <- list(hc_plot, mef_plot, hc_ndvi_plot, mef_ndvi_plot)

(hypontwintfig <- arrangeGrob(
  grobs = gl_phen,
  #heights = unit(c(3, 3), c("in", "in")),
  #  widths = c(2, 1, 1),
  layout_matrix = rbind(c(1, 3),
                        c(2, 4))
))

ggsave("../figures/breed_phen.pdf", hypontwintfig)
ggsave("../figures/breed_phen.png", hypontwintfig)
```


# Winter distribution and seasonality

Extract winter observations and then remove duplicated entries (i.e. identical geometries)
```{r}
#filter to only winter records
ebd_wint <- ebd_sf[ebd_sf$season == "winter",]
#remove duplicated identical geometries
ebd_wint_U <- ebd_wint[!duplicated(ebd_wint$geometry),]
#still leaving some minor differences in geom for the one bird at South padre 
#island reported by a million birders.  Remove all but one manually
ebd_wint_U <- ebd_wint_U[-c(28:37),]
```

Plot worldclim seasonality with eBird winter distributions

```{r}
#winter ebird map
(ebd_wint <- ggplot() +
   #pol boundaries
   geom_sf(data = world) +
   geom_sf(data = usa_states) +
   geom_sf(data = mx_states) +

   #wordclim seasons
   layer_spatial(data=wclim_4_crop, alpha=0.8) +
   scale_fill_viridis(name = 
                        expression(paste("SD(", bar(T)[monthly], ") X 100"))) +
   
   #eBird obs
   geom_sf(data = ebd_wint_U, inherit.aes = F,
           size = 1) +
   #themes, etc.
   coord_sf(xlim = c(-110, -90), ylim = c(10, 40), expand = F) +
   #geom_text(aes(x=-110, y = 50, label = "eBird Winter Observations")) +
   theme_few() +
   theme(legend.position = "bottom") +
   xlab("Longitude") +
   ylab("Latitude"))

ggsave("../figures/winter_worldclim_ebd.png")
```

Get worldclim value for each observation.

```{r}
season <- raster::extract(x = wclim_4_crop, y = ebd_wint_U)

y <- density(na.omit(season), n = 2^12)
wint_season <- data.frame(season = y$x, dens = y$y)

```

Density plot

```{r}
ggplot(wint_season) +
  geom_segment(aes(x = season, y = dens, xend = season, yend = 0, color = season))+ 
  scale_fill_viridis(name = expression(paste("SD(", bar(T)[monthly], ") X 100")),
                     limits = c(0,1200)) +
  theme_few() +
  theme(legend.position = "none")
```

